Word Identification Under Multimodal Uncertainty

### 2017-02-02 write-up/main.tex
text tweaks, length

This multi-modal integration task takes place in a noisy world. On the auditory side, individual acoustic word tokens are almost always ambiguous with respect to the particular sequence of phonemes they represent, which is due to the inherent variability of how a phonetic category is realized acoustically \cite<e.g.,>{hillenbrand1995}. And some tokens may be distorted additionally by mispronunciation or ambient noise. Perhaps the speaker was yelling ``pea" and not ``bee.'' Similarly, a sensory impression may not be enough to make a definitive identification of a visual category.\footnote{In the general case, language can of course be visual as well as auditory, and object identification can be done through many modalities. For simplicity, we focus on audio-visual matching here.} Perhaps the quickly flying insect was a beetle or a fly instead.

@@

In this paper, we propose a probabilistic framework where such reasoning can be expressed in a precise way. We characterize uncertainty, in each modality, with a probability distribution, and we predict ideal responses by combining these probabilities in a optimal way. Our work can be seen as an extension to previous Bayesian models of phoneme identification \cite<e.g.,>{feldman2009}, where, instead of a unimodal input, we model a bimodal one. A few studies have explored some aspects of audio-visual processing in a probabilistic framework  \cite<e.g.,>{bejjanki2011, kleinschmidt2015}. In these studies, the researchers focused on the specific case of phoneme recognition from speech and lip movement, however, where information from both modalities is tightly correlated.

In the present work, we study, rather, the case of arbitrary mapping between sounds and visual objects. We test participants on their ability to process audio-visual stimuli and use them to recognize the underlying word. More precisely we study the case where both the word's form and the word's referent are ambiguous, and we examine the extent to which humans conform to, or deviate from the predictions of the ideal observer model. Moreover, since some previous studies on audio-visual processing documented cases of modality preference (i.e., when people rely predominantly on the visual modality \cite<e.g., >{colavita74} or the auditory modality \cite<e.g., >{sloutsky2003}), we will  explore if, in our task as well, people show evidence of a modality preference.

The paper is organized as follows. First, we introduce our audio-visual recognition task. We next present the ideal observer model. Then we present two behavioral experiments where we test word recognition under audio-visual uncertainty. In Experiment 1, audio-visual tokens are ambiguous with respect to their category membership. In Experiment 2, we intervene by adding  noise to one modality. In both experiments participants show qualitative patterns of optimal behaviour. Moreover, while participants show no modality preference in Experiment 1, in Experiment 2 they over-rely on visual input when the auditory modality is noisy.

@@

In each trial, participants are presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test stimulus (Figure \ref{fig:task}). The test stimulus may differ from the target in both the auditory and the visual components.  After these two presentations, participants are instructed to press ``same'' or ``different.''

@@

In the bimodal condition, participants see word tokens with audio-visual input, and have to make a categorization decision. We define word tokens as vectors in the audio-visual space, $\mathbf{w}=(a,v)$.
A word category $W$ is defined as the joint distribution of auditory and visual categories. It can be characterized with a bivariate normal distribution in the audio-visual space:
$$ p(\mathbf{w} | W) \sim  N(M_W, \Sigma_W) $$

@@

We make the assumption that, given a particular word category, the auditory and visual tokens are independently generated:

@@

 with $\beta_a=\frac{\mu_{A1}-\mu_{A2}}{\sigma^2_{A}}$,
$\beta_v=\frac{\mu_{V1}-\mu_{V2}}{\sigma^2_{V}}$,
$\beta_0=\frac{\mu^2_{A2}-\mu^2_{A1}}{2\sigma^2_{A}}+\frac{\mu^2_{V2}-\mu^2_{V1}}{2\sigma^2_{V}}$

and $1+\epsilon=\frac{p(W_1)}{p(W_2)}$ is the proportion of the prior probabilities. If the identity of word categories is randomized, and if $W_1$ is the target, then $\epsilon$ measures a response bias to ``same'' if $\epsilon > 0 $, and a response bias to ``different'' if $\epsilon < 0 $.

@@

In Experiment 1, we test the predictions of the model in the case where uncertainty is due to similar auditory categories, and similar visual categories. Crucially, the similarity is such that the distributions overlap. To simulate such uncertainty in a controlled fashion, we use a continuum along the second formant (F2) linking the words /aba/ and /ada/, and we use a morph that links the prototypical pictures of the dog and the cat.

@@

The first part of the experiment trained participants using only the prototype pictures and the prototype sounds (12 trials, 4 each from the bimodal, audio-only, and visual-only conditions). After completing training, we instructed participants on the structure of the task and encouraged them to base their answers on both the sounds and the pictures (in the bimodal condition). There were a total of 25 possible combinations in the bimodal condition, and 5 in each of the unimodal conditions. Each participant saw each possible trial in each condition twice, for a total of 70 trials/participants. Trials were blocked by condition and blocks were presented in random order.
% We designed the experiment to take between 10 to 15 minutes to complete.

\begin{figure}
\centering
\includegraphics[width=3.25in\textwidth]{unimodal.png}
\caption{Average human responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\% confidence intervals. Solid lines represent unimodal logistic fits.}
\label{fig:unimodal}
\end{figure}

@@

with $\beta_a=\frac{\mu_{A_1}-\mu_{A_2}}{\sigma^2_{A}}$ and  $\beta_{a0}=\frac{\mu^2_{A_2}-\mu^2_{A_1}}{2\sigma^2_{A}}$. $\epsilon_A$ is the response bias in the auditory-only trials.

@@

Qualitatively, participants' judgments were similar to the predictions of the ideal observer model (remember that the latter was obtained by optimally combining fits to the unimodal data). Consider, for example, the contrast between the auditory-only case (dashed black line in Figure \ref{fig:Exp}, top left) and the bimodal case (solid colored lines).  Higher certainty in the visual modality generally influenced responses, with greater visual distance leading to more ``different'' ratings and less visual distance leading to more ``same'' ratings. Similar observations can be made about the contrast between the visual-only case and the bimodal case.  Overall, we found that the ideal observer model explained almost 90\% of the variance in judgments ($r^2=0.89$).

But although we see a qualitative resemblance between human data and the ideal observer model, there were quantitative differences. For example, model predictions were more influenced by the visual modality at the auditory midpoint (the point of highest uncertainty) than human judgements, and were more compressed at the endpoints (the points of lowest uncertainty).

Formally speaking, there was an increase in the value of the variance associated with each modality. More precisely, whereas the ideal observer model predicted the weights to be proportional to $1/\sigma^2_{A}$ and $1/\sigma^2_{V}$, for the auditory and the visual modalities, respectively (see expression \ref{eq:pred}), the fit to human data suggested that the real weights were proportional to $1/\sigma^2_{Ab}$ and $1/\sigma^2_{Vb}$. Interestingly, our analysis of modality preference showed that the relative values of these variances (i.e., $\frac{\sigma^2_V}{\sigma^2_A}$ vs. $\frac{\sigma^2_{Vb}}{\sigma^2_{Ab}}$) were almost the same (Figure \ref{fig:pref}, left). This means two things: 1) the bimodal presentation caused reliance on both modalities to decrease, and 2) this decrease did not affect the relative weighting of both modalities, i.e., participants were weighting modalities according to their relative reliability. The latter explains the qualitative resemblance between the predictions of the ideal observer and human data, and the former explains the quantitative discrepancy.
\newline

In Experiment 1, we tested the ability of people to recognize a a word under audio-visual uncertainty linked to category membership. We found that, generally speaking, participants followed the ideal observer model in that they weighted modalities according to their relative reliability. In real life situations, however, tokens can undergo distortions due to noisy factors in the environment. In Experiment 2, we try to account for this additional level of uncertainty.

@@

Participants' decision threshold suggested a preference for the visual modality (the non-noisy modality). Indeed non-parametric resampling of the data showed there to be a decrease in the value of the slope (\ref{fig:pref}, right).

\subsubsection{Discussion}

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{preference.png}
\caption{Top: decision thresholds in the audio-visual space. Red dotted line is the prediction of the ideal observer. Blue dotted lines are cases where modality preference is twice as strong as the ideal observer. Solid line is the threshold derived from human data. Bottom: comparison of the threshold slope between the ideal observer and the fit to human data. Error bars are 95\% confidence intervals computed through non-parametric bootstrap sampling (10.000 samples)}
\label{fig:pref}
\end{figure}

We found, similar to Experiment 1, that participants generally showed qualitative patterns similar to the ideal observer model ($r^2 = .91$). But we also found a similar discrepancy at the quantitative level, suggesting an effect of bimodal presentation on performance. Formally speaking, the ideal observer model predicted the modality weights to be proportional to $1/(\sigma^2_{A}+\sigma^2_{N})$ and $1/\sigma^2_{V}$, for the auditory and the visual modalities, respectively. The fit to human data suggested that the empirical weights were proportional to $1/\sigma^2_{Ab}$ and $1/\sigma^2_{Vb}$. An interesting difference with Experiment 1, however, was that participants had a clear preference for the non-noisy modality, as the values of the relative variances $\frac{\sigma^2_V}{\sigma^2_A + \sigma^2_N}$ vs. $\frac{\sigma^2_{Vb}}{\sigma^2_{Ab}}$ were different (\ref{fig:pref}, right). This preference affected the relative weighting, where, contrary to Experiment 1, the visual modality now has a greater weight than what could be expected from the relative reliability alone.

@@

\section{General discussion}

@@

There were, however, quantitative differences: the audio-visual presentation caused the reliance on both modalities to decrease to a lower extent than the ideal observer model expected. This phenomenon could be attributed to the arbitrary nature of the form-meaning mapping. Previous studies suggest, indeed, that while redundant multi-modal information improves performance (e.g., determining the frequency of a bouncing ball from visual and auditory cues), arbitrary mappings generally tends to hinder performance \cite<for review, see>{robinson2010}. Another possibility could be that the ideal model failed to capture some other features of multi-modal integration. Further work is needed to understand this result.

Interestingly, however, in Experiment 1 this decrease in reliance (or the increase in variance) occurred at a similar rate for both the auditory and the visual modality, and thus, it did not affect their relative weighting. The latter was primarily determined by informational reliability. Only when we intervened by adding  noise to one modality in Experiment 2, did participants show a systematic preference for the non-noisy modality. One possible explanation for this preference could be that people do not combine cross-modal uncertainties of a similar kind (e.g., category ambiguity in both modalities) in the same way they would combine uncertainties of different kinds (e.g., category ambiguity in one modality and noise in the other). Further work is needed to determine the extent to which this is the case.

Overall, in both Experiments, the majority of the variance (around 90\%) could be explained by an ideal observer that combined multi-modal information in an optimal way. In the light of this main results, we can revisit some previous findings in the literature. For instance, \citeA{sloutsky2003} reported a dominance for the auditory modality in children. This dominance disappear or reverses in adults. Could this developmental difference be driven by changes in the intrinsic relative reliability of modalities? More work is needed to carefully examine this (and other) speculations, and more generally, to determine the extent to which the optimal combination account helps us better understand the mechanisms of word precessing and learning.

%However, is it the case that children weight modalities differently because a sound is inherently more salient than any visual contrast? or could part of this effect stem from the fact that the precise sound and visual contrasts used as stimuli differ in their relative salience to children? In particular, what happens if the sound contrast is reduced to a perceptual minimum and when the visual contrast is enhanced? Can we reverse the weighting in this case as our optimal combination account would suggest? Moreover, could the developmental difference be driven by changes in the relative reliability of modalities?

%In a different line of work, it has been shown that infants have difficulties mapping minimally different words to two novel objects, despite the fact that they can perceptually differentiate these words \cite<e.g.,>{stager1997}. Is it possible that what had been interpreted as `failure' in learning could, at least to some extent, be explained by a categorical decision (similar to what adults performed in our task). More precisely, it is possible that the visual stimuli were quite similar in the children’s early conceptual space, which made them believe that both objects were possible instances of the same semantic category. This, in turn, would make them interpret the sound difference as alternative pronunciations of the same word. In particular, what happens if the object contrast is maximized? More work is needed to carefully examine these (and other) speculations, and more generally, to determine the extent to which the optimal combination account helps us better understand the mechanisms of word precessing and learning.

>>>>

This multi-modal integration task takes place in a noisy world. On the auditory side, individual acoustic word tokens are almost always ambiguous with respect to the particular sequence of phonemes they represent, which is due to the inherent variability of how a phonetic category is realized acoustically \cite<e.g.,>{hillenbrand1995}. And some tokens may be distorted additionally by mispronunciation or ambient noise. Perhaps the speaker was yelling ``pea" and not ``bee.'' Similarly, a sensory impression may not be enough to make a definitive identification of a visual category.\footnote{In the general case, language can of course be visual as well as auditory, and object identification can be done through many modalities. For simplicity, we focus on audio-visual matching here.} Perhaps the insect was a beetle or a fly instead.

@@

In this paper, we propose a probabilistic framework where such reasoning can be expressed precisely. We characterize uncertainty in each modality with a probability distribution, and we predict ideal responses by combining these probabilities in a optimal way. Our work can be seen as an extension to previous Bayesian models of phoneme identification \cite<e.g.,>{feldman2009}, where, instead of a unimodal input, we model a bimodal one. A few studies have explored some aspects of audio-visual processing in a probabilistic framework  \cite{bejjanki2011, kleinschmidt2015}. In these studies, the researchers focused on the specific case of phoneme recognition from speech and lip movement, however, where information is tightly correlated across modalities.

In the present work, we study, rather, the case of arbitrary mapping between sounds and visual objects. We test participants on their ability to process audio-visual stimuli and use them to recognize the underlying word. More precisely we study the case where both the word's form and the word's referent are ambiguous, and we examine the extent to which humans conform to, or deviate from the predictions of the ideal observer model. Moreover, some previous studies on audio-visual processing documented cases of modality preference, when people rely predominantly on the visual modality \cite<e.g., >{colavita1974} or the auditory modality \cite<e.g., >{sloutsky2003}. Thus, we will explore if participants in our task show evidence of a modality preference.

The paper is organized as follows. First, we introduce our audio-visual recognition task. We next present the ideal observer model. Then we present two behavioral experiments where we test word recognition under audio-visual uncertainty. In Experiment 1, audio-visual tokens are ambiguous with respect to their category membership. In Experiment 2, we intervene by adding  noise to one modality. In both experiments participants show qualitative patterns of optimal behavior. Moreover, while participants show no modality preference in Experiment 1, in Experiment 2 they over-rely on visual input when the auditory modality is noisy.

@@

In each trial, participants are presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test stimulus (Figure \ref{fig:task}). The test stimulus may differ from the target in both the auditory and the visual components.  After these two presentations, participants press ``same'' or ``different.''

@@

In the bimodal condition, participants see word tokens with audio-visual input, and have to make a categorization decision. We define word tokens as vectors in the audio-visual space, $\mathbf{w}=(a,v)$.
A word category $W$ is defined as the joint distribution of auditory and visual categories. It can be characterized with a bivariate normal distribution:
$$ p(\mathbf{w} | W) \sim  N(M_W, \Sigma_W) $$

@@

We make the assumption that, given a particular word category, the auditory and visual tokens are independent:

@@

\noindent with $\beta_a=\frac{\mu_{A1}-\mu_{A2}}{\sigma^2_{A}}$,
$\beta_v=\frac{\mu_{V1}-\mu_{V2}}{\sigma^2_{V}}$,
$\beta_0=\frac{\mu^2_{A2}-\mu^2_{A1}}{2\sigma^2_{A}}+\frac{\mu^2_{V2}-\mu^2_{V1}}{2\sigma^2_{V}}$
and $1+\epsilon=\frac{p(W_1)}{p(W_2)}$ is the proportion of the prior probabilities. If the identity of word categories is randomized, and if $W_1$ is the target, then $\epsilon$ measures a response bias to ``same'' if $\epsilon > 0 $, and a response bias to ``different'' if $\epsilon < 0 $.

@@

In Experiment 1, we test the predictions of the model in the case where uncertainty is due to similar auditory categories, and similar visual categories. Crucially, the similarity is such that the distributions overlap. To simulate such uncertainty in a controlled fashion, we use a continuum along the second formant (F2) linking the words /aba/ and /ada/, and we use a morph that links dogs and cats.

@@

The first part of the experiment trained participants using only the prototype pictures and the prototype sounds (12 trials, 4 each from the bimodal, audio-only, and visual-only conditions). After completing training, we instructed participants on the structure of the task and encouraged them to base their answers on both the sounds and the pictures (in the bimodal condition). There were a total of 25 possible combinations in the bimodal condition, and 5 in each of the unimodal conditions. Each participant saw each possible trial twice, for a total of 70 trials/participant. Trials were blocked by condition and blocks were presented in random order.
% We designed the experiment to take between 10 to 15 minutes to complete.

\begin{figure}
\centering
\includegraphics[width=3.25in]{unimodal.png}
\caption{Average human responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\% confidence intervals. Solid lines represent unimodal logistic fits.}
\label{fig:unimodal}
\end{figure}

@@

\noindent with $\beta_a=\frac{\mu_{A_1}-\mu_{A_2}}{\sigma^2_{A}}$ and  $\beta_{a0}=\frac{\mu^2_{A_2}-\mu^2_{A_1}}{2\sigma^2_{A}}$. $\epsilon_A$ is the response bias in the auditory-only trials.

@@

Qualitatively, participants' judgments were similar to the predictions of the ideal observer model (remember that the latter was obtained by optimally combining fits to the unimodal data). Consider, for example, the contrast between the auditory-only case (dashed black line in Figure \ref{fig:Exp}, top left) and the bimodal case (solid colored lines).  Higher certainty in the visual modality generally influenced responses, with greater visual distance leading to more ``different'' ratings and less visual distance leading to more ``same'' ratings. Similar observations can be made about the contrast between the visual-only case and the bimodal case.

Overall, we found that the ideal observer model explained much of the variance in judgments ($r^2=0.89$). But although we see a qualitative resemblance between human data and the model, there were quantitative differences. For example, model predictions were more influenced by the visual modality at the auditory midpoint (the point of highest uncertainty) than human judgements, and were more compressed at the endpoints (the points of lowest uncertainty).

Formally, there was an increase in the value of the variance associated with each modality. Whereas the ideal observer model predicted the weights to be proportional to $1/\sigma^2_{A}$ and $1/\sigma^2_{V}$, for the auditory and the visual modalities, respectively (see expression \ref{eq:pred}), the fit to human data suggested that the real weights were proportional to $1/\sigma^2_{Ab}$ and $1/\sigma^2_{Vb}$. Our analysis of modality preference showed that the relative values of these variances were almost the same (Figure \ref{fig:pref}, left). Thus, 1) the bimodal presentation caused reliance on both modalities to decrease, and 2) this decrease did not affect the relative weighting of both modalities, i.e., participants were weighting modalities according to their relative reliability. The latter explains the qualitative resemblance between the predictions of the ideal observer and human data, and the former explains the quantitative discrepancy.

% (i.e., $\frac{\sigma^2_V}{\sigma^2_A}$ vs. $\frac{\sigma^2_{Vb}}{\sigma^2_{Ab}}$)

In sum, we found that participants followed the ideal observer model in that they weighted modalities according to their reliabilities. In real life, however, tokens can undergo distortions due to noisy factors in the environment. In Experiment 2, we explore this additional level of uncertainty.

@@

Participants' decision threshold suggested a preference for the visual modality (the non-noisy modality). Indeed non-parametric resampling of the data showed a decrease in the value of the slope (\ref{fig:pref}, right).

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{preference.png}
\caption{Top: decision thresholds in the audio-visual space. Red dotted line is the prediction of the ideal observer. Blue dotted lines are cases where modality preference is twice as strong as the ideal observer. Solid line is the threshold derived from human data. Bottom: comparison of the threshold slope between the ideal observer and the fit to human data. Error bars are 95\% confidence intervals computed via non-parametric bootstrap.}
\label{fig:pref}
\end{figure}

We found, similar to Experiment 1, that participants generally showed qualitative patterns similar to the ideal observer model ($r^2 = .91$). But we also found a similar discrepancy at the quantitative level. The ideal observer model predicted the modality weights to be proportional to $1/(\sigma^2_{A}+\sigma^2_{N})$ and $1/\sigma^2_{V}$, for the auditory and the visual modalities, respectively. The fit to human data suggested that the empirical weights were proportional to $1/\sigma^2_{Ab}$ and $1/\sigma^2_{Vb}$. An interesting difference with Experiment 1, however, was that participants had a clear preference for the non-noisy modality, as the values of the relative variances were different (Figure \ref{fig:pref}, right). This preference affected the relative weighting, where, contrary to Experiment 1, the visual modality had greater weight than what could be expected from its relative reliability alone.
% $\frac{\sigma^2_V}{\sigma^2_A + \sigma^2_N}$ vs. $\frac{\sigma^2_{Vb}}{\sigma^2_{Ab}}$

@@

\section{General Discussion}

@@

There were, however, quantitative differences. Audio-visual presentation caused the reliance on both modalities to decrease to a lower extent than the ideal observer model expected. One possbible explanation is that this phenomenon was caused by the arbitrary nature of the form-meaning mapping. Previous studies suggest that while redundant multi-modal information improves performance (e.g., determining the frequency of a bouncing ball from visual and auditory cues), arbitrary mappings generally tends to hinder performance \cite<for review, see>{robinson2010}.
% Or perhaps the ideal model failed to capture some other features of multi-modal integration.
% Further work is needed to understand this result.

Interestingly, however, in Experiment 1 this decrease in reliance (or the increase in variance) occurred at a similar rate for both the auditory and the visual modality, and thus, it did not affect their relative weighting. The latter was primarily determined by informational reliability. Only when we intervened by adding  noise to one modality in Experiment 2, did participants show a systematic preference for the non-noisy modality. One possible explanation for this preference could be that people do not combine cross-modal uncertainties of a similar kind (e.g., ambiguity in both modalities) in the same way they would combine uncertainties of different kinds (e.g., ambiguity in one modality and noise in the other).
% Further work is needed to determine the extent to which this is the case.

Overall, in both Experiments, the majority of the variance could be explained by an ideal observer that combined multi-modal information optimally. In the light of this main result, we can revisit some previous findings in the literature. For instance, \citeA{sloutsky2003} reported a dominance for the auditory modality in children. This dominance disappear or reverses in adults. Could this developmental difference be driven by changes in the intrinsic relative reliability of modalities across development? More work is needed to carefully examine this (and other) speculations, and more generally, to determine the extent to which the optimal combination account helps us better understand the mechanisms of word precessing and learning.
